\fontfamily{phv}\selectfont{Causal Discovery methods for Effective Connectivity in Human Brains}

### 2014-01-16 docs/paper/paper final.tex
stuk over validation toegevoegd. Nog niet tevreden over data acquisition, dat is lastig te verwoorden

\subsection{Research goal}
%How does the standard causal discovery method PC-Algorithm behave regarding the aforementioned problems when applied to resting-state functional time-series data?
%- Can we overcome the consistency problems of PC-algorithm applied on resting-state functional data?
%- Does the PC algorithm applied on resting-state functional data provide consistent, specific and anatomically plausible connectivity?
%- How does functional connectivity found by the first part of PC relate to other structural / functional methods applied on the same resting-state functional data regarding the aforementioned problems?
%
%How does the performance of EMS-PC relate to PC regarding the aforementioned problems when applied to the same resting-state functional time-series data?
%- When would EMS-PC perform better than PC-algorithm regarding directionality results?
%- Is EMS-PC more consistent in its results within a single subject than PC-algorithm on resting-state functional data, regarding the multiple separating set addition and the explicit v-structure test?
%- Can we find unfaithfulness within our model without additional independence tests?

@@

%Before finding directed edges, the PC-algorithm first finds undirected structure.
%For each two non-adjacent nodes $X$ and $Y$ standard PC finds a minimal separating set, which is the smallest set acting as a separating set.
%This is done by first assuming all nodes are connected, and then testing all pairs of nodes for independence.

@@

\paragraph{Run-time performance}
The PC-algorithm run-time is in the order of the amount of dependency tests to be performed and thus it has a worst-case runtime complexity exponential in the maximal branching degree present in the graph.
However, this worst-case scenario only happens when no edges can be removed; all subsets of the complete set of nodes have to be used to test d-separation of every pair of nodes.
For a sparse graph, many edges may be removed for small separating sets, and this enables the standard PC algorithm to often run fast in practice perform well on hundreds of variables. % TODO: add citation to example publication. Peter Lucas maybe or examples in the sprites book?

@@

\paragraph{Multiple Separating Sets}
As will be explained in the results section, standard PC is a bit hasty in assigning v-structures, resulting in many unnecessary and perhaps unfaithful directionality.
The idea of CPC, to make a weaker assumption on graph faithfulness than standard PC is potential as less erroneous arrows are found.
Instead of doing additional tests as described in CPC, one could attempt to determine causal unfaithfulness with less dependency tests.
An adaptation to PC algorithm that would not increase the order of complexity is once a separating set is found to test all other vertex subsets of the same order as well.
This finds all separating sets of the lowest order without increasing complexity.
Orienting an unshielded triple $\left<A,B,C\right>$ requires $B$ not to occur in any separating set of $(A,C)$.
By having more than one separating set, this constraint is strengthened is the data is not faithful.
If it is, the results are not changed.
This also does not change the run-time complexity.

@@

\subsection{Experimental validation}
\paragraph{Experimental pipeline}
%- What data is needed
%- Apply structure PC on resting-state functional time-series
%- Apply directional PC
%- Parameter optimisation (model selection)

\paragraph{Standard PC}
%- Measuring whole-brain structural and directional consistency of PC - frequency of occurrences within subjects
%- Markov equivalence class size (nr. of latent variables / bidirectional arrows) within subjects
%- Markov class Asymmetry and certain asymmetry within subjects
%- Consistency of structure across subjects when aggregating results (averaging and frequency of occurences)
%- Consistency of directionality across subjects when aggregating results

\paragraph{EMS-PC}
%- Measuring whole-brain structural and directional consistency of EMS-PC
%- Markov equivalence class size (nr. of latent variables / bidirectional arrows) within subjects
%- Markov class Asymmetry and certain asymmetry within subjects
%- Amount of separating sets found
%- Unfaithfulness check with and without additional separating sets

\paragraph{Comparison}
%- Comparison of directionality between standard PC and EMS-PC on consistency of results within subjects
%- Comparison of ... of results across subjects
%- Run-time of methods and comparison between run-times
%- Comparison of structure (functional connectivity) of PC (both. they're identical in structure) with DWI results by Max et al.

\subsection{Data acquisition}
%Subject acquisition and instruction
%Used equipment and specs
%Experimental setup
%Scanner settings
%Preprocessing
%- Software
%- Scripts
%Resulting data specs
%[Ramon: tot hoeverre is dit in detail nodig? Enerzijds moeten we niet doen alsof het kant en klaar uit dat apparaat komt, maar anderzijds wordt het in Max' artikel beschreven. Zijn dingen als "settings" dan niet overkill?]

>>>>

\subsection{Research goal}
%How does the standard causal discovery method PC Algorithm behave regarding the aforementioned problems when applied to resting-state functional time-series data?
%- Can we overcome the consistency problems of PC algorithm applied on resting-state functional data?
%- Does the PC algorithm applied on resting-state functional data provide consistent, specific and anatomically plausible connectivity?
%- How does functional connectivity found by the first part of PC relate to other structural / functional methods applied on the same resting-state functional data regarding the aforementioned problems?
%
%How does the performance of EMS-PC relate to PC regarding the aforementioned problems when applied to the same resting-state functional time-series data?
%- When would EMS-PC perform better than PC algorithm regarding directionality results?
%- Is EMS-PC more consistent in its results within a single subject than PC algorithm on resting-state functional data, regarding the multiple separating set addition and the explicit v-structure test?
%- Can we find unfaithfulness within our model without additional independence tests?

@@

%Before finding directed edges, the PC algorithm first finds undirected structure.
%For each two non-adjacent nodes $X$ and $Y$ standard PC finds a minimal separating set, which is the smallest set acting as a separating set.
%This is done by first assuming all nodes are connected, and then testing all pairs of nodes for independence.

@@

\paragraph{Run-time performance}
The PC algorithm run-time is in the order of the amount of dependency tests to be performed and thus it has a worst-case runtime complexity exponential in the maximal branching degree present in the graph.
However, this worst-case scenario only happens when no edges can be removed; all subsets of the complete set of nodes have to be used to test d-separation of every pair of nodes.
For a sparse graph, many edges may be removed for small separating sets, and this enables the standard PC algorithm to often run fast in practice perform well on hundreds of variables. % TODO: add citation to example publication. Peter Lucas maybe or examples in the sprites book?

@@

\paragraph{Multiple Separating Sets}
As will be explained in the results section, standard PC is a bit hasty in assigning v-structures, resulting in much unnecessary and perhaps unfaithful directionality.
The idea of CPC, to make a weaker assumption on graph faithfulness than standard PC is potential as less erroneous arrows are found.
Instead of doing additional tests as described in CPC, one could attempt to determine causal unfaithfulness with less dependency tests.
An adaptation to PC algorithm that would not increase the order of complexity is once a separating set is found to test all other vertex subsets of the same order as well.
This finds all separating sets of the lowest order without increasing complexity.
Orienting an unshielded triple $\left<A,B,C\right>$ requires $B$ not to occur in any separating set of $(A,C)$.
By having more than one separating set, this constraint is strengthened if the data is not faithful.
If it is, then the results are not changed.
This also does not change the run-time complexity.

@@

%stability and iterations TODO: right place for this? I think it is, as it can be seen as a simple extension to naive PC
\paragraph{Stability}
The graph resulting from the PC algorithm depends on the order of the nodes\cite[p.120-122]{spirtes2000}.
It is therefore questionable whether strong conclusions about individual edges can be drawn from the output, as such an edge might not have appeared if the input was ordered differently.
As such, it seems useful to calculate a chance of a connection appearing.
When running the experiments, we have applied the algorithms multiple times using random permutations of the nodes to obtain stability selection as described by Meinshausen and B{\"u}hlmann \cite{meinshausen2010}.
These results have been avaraged to conclude the certainty of the found edges.

\subsection{Experimental validation}
\paragraph{Experimental pipeline}{sec:pipeline}
%- What data is needed
%- Apply structure PC on resting-state functional time-series
%- Apply directional PC
Both standard PC and EMS-pc have been used to find brain connectivity.
The data which is used is functional time-series data, and it is the same as the data used in Hinne et al.\cite{hinne2013}, as described in Section~\ref{sec:data}.
This data is obtained from six subject.
From this data, covariance between regions can be determined, so that regions can be tested for \emph{d}-separation with the Fisher's conditional independence test.
This is required to use the PC algorithm.
First, the strucural part of the PC algorithm will be applied on the data.
This will result in a graph describing undirected connections between regions.
There is no difference between standard PC and EMS-PC in this result.
After that, the second part of the PC algorithm is run to orient edges in this graph.
This is different for standard PC and EMS-PC and a result will therefore be obtained for both algorithms.
%- Parameter optimisation (model selection)
The experiments were run using an alpha-value of 0.01 for the independence test, and using 300 random permutations of the nodes per subject.

\paragraph{Concluding connectivity from PC}
%- Measuring whole-brain structural and directional consistency of PC - frequency of occurrences within subjects
Applying the standard PC algorithm yields an undirected graph, in which each edge has a certainty.
This is done both per subject, and as an avarage over all subjects.
We compared the generated structure with a Bayesian inference approach on finding structural connectivity from diffusion weighted imaging, as described in Hinne et al.\cite{hinne2013}.

In the second stage of PC, a directed graph is found.
Again, a certainty of each individual arrowhead may be found by applying stability selection.
%- Markov equivalence class size (nr. of latent variables / bidirectional arrows) within subjects
Different permutations of nodes may yield different directionalities for the same edge, and bi-directional arrows may be found.
Of specific interest here are consistent non-symmetrical orientations, as one-directional arrows imlpy a causal relation.
Edges of two arrowhead imply a common confounder.
%- Markov class Asymmetry and certain asymmetry within subjects
%- Consistency of structure across subjects when aggregating results (averaging and frequency of occurences)
The results will again be analyses per subject, and as an avarage over all subjects; Causal relations appearing in specific subjects will, or in general.
%- Consistency of directionality across subjects when aggregating results
The functioning of the brain will not completely differ between different subjects and therefore some consistency across subjects is expected.
Consistency between the two hemispheres is also expected.

%\paragraph{EMS-PC}   Er werd veel herhaald t.o.v. standard PC, dus ik heb ze samen genomen. amount of sepsets en faithfulness is niet dubbel, maar hebben we deze resultaten opgeslagen en kunnen we er iets nuttigs over zeggen?
%- Measuring whole-brain structural and directional consistency of EMS-PC
%- Markov equivalence class size (nr. of latent variables / bidirectional arrows) within subjects
%- Markov class Asymmetry and certain asymmetry within subjects
%- Amount of separating sets found
%- Unfaithfulness check with and without additional separating sets

\paragraph{Comparison between standard PC and EMS-PC}
%- Comparison of directionality between standard PC and EMS-PC on consistency of results within subjects
As EMS-PC is more constrained in concluding directionality, it may find less erroneous arrows than standard PC.
To verify this, the number and consistency of one-directional arrows is compared.
%- Comparison of ... of results across subjects
Again, this is both done per-subject and on the average across subjects.
%- Run-time of methods and comparison between run-times
The run-time of both methods will be compared.
%- Comparison of structure (functional connectivity) of PC (both. they're identical in structure) with DWI results by Max et al. -> verplaatst naar het stukje aan het begin, over structure PC

\subsection{Data acquisition}\label{sec:data}
%Subject acquisition and instruction
%Used equipment and specs
%Experimental setup
%Scanner settings
%Preprocessing
%- Software
%- Scripts
%Resulting data specs
%TODO niet helemaal happy, hoe zeg je dit zonder steeds "as described by Hinne et al." te zeggen? "This research" kan ook naar ons eigen werk verwijzen en is verwarrend, het moet duidelijk zijn dat dit alles niet door ons gedaan is.
We performed our analysis on the same data as treated by Hinne et al. in the aforementioned Bayesian structural analysis \cite{hinne2013, hinne2013structfunc}.
This article describes how resting-state functional data of six subjects was measured at the Donders Centre for Cognitive Neuroimaging at the Radboud University.
This functional data is divided in 119 nodes and 1029 time-series per subject.
Additionaly, diffusion-weighted images of these subjects have been obtained.
Based on these DWI-measurements, a structural connectivity matrix has been inferred with a technique described by Hinne et al.
We used the six resulting resting-state functional time series data and connectivity matrices from the Bayesian analysis as pre-processed by Hinne et al.

No additional pre-processing has been done in our research in order to apply the PC algorithm on this data.
